Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-05-23 paper_nips/paper.tex
methods standing

%
\section{Methods}
%
Using
Human Connectome Project (HCP) data (n=500), optimal low-rank projections and
logistic-regression models are identified in a same gradient descent. Brain
network decompositions are thus exposed that explain task-descriminative spatial
patterns.

Data was drawn from 498 unrelated, healthy HCP participants.
All provided informed consent to the Washington
University in St. Louis institutional review board. HCP tasks 
were selected that feature known suitability as localizers
and reliability
across participants \cite{barch2013}.
Mostly block-design, but also event-related, paradigms were
administered on 1) working memory/cognitive control processing, 2)
incentive processing, 3) visual and somatosensory-motor processing,
4) language processing (semantic and phonological processing),
5) social cognition, 6) relational processing, and 7) emotional
processing. All data were acquired on the same Siemens Skyra 3T scanner.
Whole-brain EPI acquisitions were acquired with a
32 channel head coil (TR=720ms, TE=33.1ms, flip angle=52°, BW=2290Hz/Px,
in-plane FOV=$280\times180$mm, 72 slices, 2.0mm isotropic voxels).
The “minimally preprocessed” pipeline \cite{glass13} includes
gradient unwarping, motion correction, fieldmap-based EPI distortion
correction, brain-boundary-based registration of EPI to structural
T1-weighted scan, non-linear (FNIRT) registration into MNI space,
and grand-mean intensity normalization. Activation maps were spatially
smoothed by a Gaussian kernel of 4mm (FWHM). A general linear model (GLM) was
implemented by FILM from the FSL suite with model regressors from convolution
with a “canonical” hemodynamic response function and from temporal derivatives.
HCP tasks were conceived to modulate activation
in a maximum of different brain regions and neural systems. Indeed, at
least 70\% of the participants showed consistent brain activity in
contrasts from the task battery, which certifies excellent
coverage \cite{barch2013}.
In sum, the HCP task dataset incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants.
All maps were downsampled to a common 36x43x36 space of
5mm isotropic voxels and gray-matter masked (at least 10\%).
All analyses were based on task maps of
13,657 voxels representing Z values in gray matter.
\linebreak

Unsupervised and supervised learning were combined into a low-rank logistic
regression problem. The 13,657 z values from each activity map were
subject to a first linear
projection into Z latent components (i.e., 1, 5, 10, .., 100).
These hidden brain networks
loadings were subsequently projected into the 18 class space for multinomial
logistic regression. The goal was to find the two weight matrices
(input-to-hidden: 13,657 x Z and hidden-to-output: Z x 18) and their
corresponding bias vectors. Non-linearities were not applied on the
transformation results. Weights and biases were initialized by Gaussian
random noise. Gradient descent updated these matrices and vectors in each
iteration (100 samples per batch, 250 epochs). Using the chain rule, the
partial
derivates for the update were computed for the transformation into the
latent space and the subsequent transformation into the class space. We choose
the RMSprop algorithm \cite{rmsprop} with an initial learning rate of
0.01. Early stopping ensured that the best weight matrices/vectors were
retained in each epoch cycle. This was evaluated by the prediction
accuracy on a validation set (10\% of the training data) at each iteration.
\linebreak

This approach was benchmarked against independent dimensionality reduction and
learning of a classification function. Data reduction was performed on one
half of the data by ICA and SPCA.
ICA unmixed the
BOLD signals into separate spatial components
by minimizing their mutual information \cite{hyvarinen2000}.
This iterative blind source separation was
realized by a parallel FASTICA implementation (200 maximum iterations,
per-iteration tolerance of 0.0001,
initialized by a random mixing matrix, preliminary whitening).
SPCA separated the BOLD signals into
network components with few regions, which scales well to 
large datasets \cite{varoqu2011}.
This regression-type optimization problem constrained by
$\ell_1$-penalty in an implementation
without orthogonality assumptions
(1000 maximum iterations, per-iteration tolerance of 1 * 10\textsuperscript{-8}, sparsity
alpha=1, ridge-shrinkage at 0.01, Lasso path computed with coordinate
descent). Each linear decomposition revealed the specified number of
latent network components in one half the HCP data.
\linebreak
The extracted hidden network components were subsequently used
to reduce the remain half of task maps to a considerably smaller number of component
loadings. 13,657 voxels were thus condensed into 1, 5, 10, .., 100
measures of network involvement using ridge regression
(regularization alpha parameter=0.001, using
Cholesky solver). Multinomial logistic regression was finally performed on the
brain network loadings to learn classifying the 18 cognitive tasks.
Importantly, in all approaches, the final classification models were tested
on an unseen test set (10\% of data).
\linebreak

linear decomposition + elastic net penalty -> has characteristics of a
sparse PCA
\begin{wrapfigure}{r}{0.40\textwidth}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{figures/figure3.png}
  \end{center}
  \caption {\textbf{Equilibrium between autoencoder and low-rank logistic regression}
  One learned decomposition component (out of 20) between the only-autoencoder
  (\textit{uppermost panel}) and only-logistic-regression
  (\textit{lowermost panel}) scenarios.
  The congruent structure (\textit{red, middle column})
  in the
  posterior cingulate cortex, posterior mid-cingulate cortex, and medial
  prefrontal cortex that is hence important in decompositions of the rest and
  task neuroimaging data. With increasing weight on the supervised learning,
  non-congruent structure emerges in the early (\textit{blue}) and
  lateral (\textit{red}) visual cortex
  (\textit{right column}). Matrix weights were z-scored.
  }
\end{wrapfigure}

Learning of the identity function as trivial solution is avoided by
the both bottleneck and the sparsity-enforcing ElasticNet penalty.

o maximize intranetwork homoge- neity and between-network heterogeneity.

AE:
coordinate system for points on the manifold
low-dimensional bottleneck layer

FORMULA
%X e R^d real-valued input
'reconstruction error' criterion
encoder: input -> hidden reprsentation
%Its parameter set is θ = {W, b}, where W is a d′ × d weight matrix and b is an offset vector of dimensionality d′.
%decoder: gθ′ (y) = s(W′y + b′), (3)
%with appropriately sized parameters θ′ = {W′,b′}.
where b0 and b1 are biase vectors
average reconstruction error
pen. avoids learning trivial soutoin - identitiy

coompute the error derivates of the weights
gradually modified according to possible compression algorithms

affine autoencoder with squared error loss
%tied weights beteen W0 and W1, i.e., W1 = W0^T
X hat is a dterministic function of X
It can thus be said that training an autoencoder to minimize reconstruction error amounts to maximizing a lower bound on the mutual information between input X and learnt repre- sentation Y .

rest data is already compressed to a sparse PCA remote form of 
pretraining of the first layer
bottleneck -> unter-complete representing, i.e. lossy compression

as a pretraining strategy

are decomposed into d networks,
in a way that maximizes the homogeneity of
function within each network while maximizing the heterogeneity between them.

output is vector-valued -> generative properties?
optimal transformation matrix

forward propegation: input through architecture to generate output
reconstruction cost
backward propegation: generate the deltas for all units / model paramters
corresponds to the derivative of the cost function with respect to the parameters

100 components close to sparse over-complete representation because
Xrest contains only 20 components

whereas elasticnet on the model paramters is regulization

FORMULA

\begin{eqnarray}
  \begin{split}
{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) &= \lambda\frac{1}{N_{X_{task}}} \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|x^{(i)}; W,b)) \hspace{0.2cm} \\
&+ \hspace{0.2cm} (1-\lambda)\frac{1}{N_{X_{rest}}}\|X_{rest}-\hat X_{rest}(W,b)\|^2
  \end{split}
  \label{eq:loss_equ}
\end{eqnarray}

mutiple output units
aCcording to equation \eqref{eq:loss_equ}
As solver, we chose \textit{rmsprop} \cite{rmsprop},
a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by RMSprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to be zero.

With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by

@@

minimizing the discrepancy between the orig- inal data and its reconstruction.
The required gradients are easily obtained by using the chain rule to
backpropagate error derivatives first through the decoder network
and then through the encoder network

theanopython
The analyses were performed in Python.
We used \textit{nilearn} to handle
the high-dimensional neuroimaging data 
\cite{abrah14}
and
\textit{theano} for automatic
differentiation of symbolic computation graphs
\cite{bastien2012theano, bergstra2010theano}.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
\url{http://github.com/banilo/nips2015}.

@@

@@

>>>>

%
\section{Methods}
%
\paragraph{Data.}
Unsupervised projection models into a lower-dimensional space and
supervised logistic-regression models are learned in concert
in a same parameter gradient descent.
Brain
network decompositions are thus exposed that explain task-descriminative spatial
patterns of neural activity.

As the currently biggest openly-accessible reference dataset,
we chose the Human Connectome Project (HCP) for the present analyses
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500 unrelated,
healthy HCP participants.
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants.
The task paradigms include
1) working memory/cognitive control processing, 2)
incentive processing, 3) visual and somatosensory-motor processing,
4) language processing (semantic and phonological processing),
5) social cognition, 6) relational processing, and 7) emotional
processing. All data were acquired on the same Siemens Skyra 3T scanner.
Whole-brain EPI acquisitions were acquired with a
32 channel head coil (TR=720ms, TE=33.1ms, flip angle=52°, BW=2290Hz/Px,
in-plane FOV=$280\times180$mm, 72 slices, 2.0mm isotropic voxels).
The “minimally preprocessed” pipeline includes
gradient unwarping, motion correction, fieldmap-based EPI distortion
correction, brain-boundary-based registration of EPI to structural
T1-weighted scans, non-linear (FNIRT) registration into MNI space,
and grand-mean intensity normalization. Activation maps were spatially
smoothed by a Gaussian kernel of 4mm (FWHM). A GLM was
implemented by FILM from the FSL suite with model regressors from convolution
with a “canonical” hemodynamic response function and from temporal derivatives.
HCP tasks were conceived to modulate activation
in a maximum of different brain regions and neural systems. Indeed, at
least 70\% of the participants showed consistent brain activity in
contrasts from the task battery, which certifies excellent
coverage \cite{barch2013}.
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common 60x72x60 space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
All supservised analyses were based on labeled HCP task maps of
79,941 voxels representing Z values in gray matter.

These labeled data were complemented by unlabeled activity maps
from HCP acquisitions of unconstrained resting-state activity.
These reflect brain activity in the absence of controlled thought.
In line with the goal of the present study, acquisition of the data was
specifically aimed at the study of task-rest correspondence.
From each included participant, we included two
time-series for left and right phase encoding
with 1,200 maps of multiband, gradient-echo planar imaging acquired
during a period of 15min (TR=720 ms, TE=33.1 ms, flip angle=52°,
FOV=$280\times180$mm, and 2.0mm isotropic voxels). Besides run duration,
the task acquisitions were identical to the resting-state fMRI acquisitions
for maximal compatibility between task and rest data.
We here drew on ``minimally preprocessed'' rest data
from 200 randomly selected healthy, unrelated participants.
PCA was applied to each set of 1,200 rest maps for
denoising by keeping only the 20 main modes of
variation.
In sum, the HCP rest data concatenated
8000 unlabeled, noise-cleaned rest maps with
40 brain images from each of 200 randomly selected participants.

We were further interested in the utility of the optimal low-rank projection
in one task dataset for dimensionality reduction in another task dataset.
To this end, the HCP-derived network decompositions were used as preliminary
step in the classifcation problem of another large-cohort dataset.
The ARCHI dataset \cite{pinel07} provides activity maps from
diverse experimental tasks, including auditory and visual perception, motor action,
reading, language comprehension and mental calculation.
81 right-handed healthy participants
(3 not included in present analyses due to incomplete data)
without psychiatric or
neurological history participated in four fMRI sessions acquired under
different experimental paradigms.
The functional maps were warped into
the MNI space and resampled to isotropic 3mm resolution.
Whole-brain EPI data were acquired with the same Siemens Trio with a 32
channel head coil (TR=2400ms, TE=30ms, flip angle=60°, in-plane
FOV=$19.2\times19.2$cm, 40 slices, 3.0mm isotropic voxels).
Standard preprocessing was performed with Nipype \cite{gorgo11}, including
slice timing, motion correction, alignment, and spatial normalization.
Activation maps were spatially smoothed by
a Gaussian kernel of 5mm (FWHM).
Analogous to HCP data, the second task dataset incorporated 1404
labeled, grey-matter masked, and z-scored activity maps
from 18 diverse tasks acquired in 78 participants.

\paragraph{Unsupervised layer.}
The labeled and unlabeled data were fed into a linear statistical model.
The model is composed of an autoencoder and a low-rank logistic regression.
On the one hand, the affine autoencoder takes the input 
$x$ and projects it into coordinates of a latent representation $z$
by

\begin{eqnarray}
  \begin{split}
  z &= W_0 x + b_{W_0} \\
  x' &= W_1 z + b_{W_1}
  \end{split}
  \label{eq:autoenc}
\end{eqnarray}

where $x \in \mathbb{R}^{d}$ denotes the vector of $d=79,941$ voxel from each
rest map, $z \in \mathbb{R}^{n}$ is the $n$ hidden dimensions (i.e.,
distributed macroscopical activity patterns), and 
$x' \in \mathbb{R}^{d}$ is the reconstruction vector of the original activity map
from the hidden variables. Further, ${W_0}$ denotes the weight matrix that
transforms
from input space into the hidden space (i.e., encoder),
${W_1}$ is the weight matirx for back-projection from the hidden space to the
output space (i.e., decoder), ${b_{W_0}}$ and ${b_{W_1}}$ are bias vectors.
Note that ${W_0}$ and ${W_1}$ are tied
$W_0$ = $W_1^T$.
The optimal model parameters ${W_0}, {b_{W_0}}, {b_{W_1}}$ are found by
minimizing the reconstruction difference by squared error

\begin{eqnarray}
  {\mathcal{L_{AE}}}(x, x') = || \x - \x' ||^2
\end{eqnarray}

This reconstruction error criterion equates with
maximizing a lower bound on the mutual information between
input and the learned representation.
Non-linearities were not applied on the
transformation results.

\paragraph{Supervised layer.}
On the other hand, such lossy compression by a low-dimensional bottleneck
is also imposed by the first layer of the low-rank
logistic regression.

\begin{eqnarray}
  \begin{split}
    P(Y=i|x, V_0,V_1,b_{V_0}, b_{V_1}) &= softmax_i(V_1 (V_0 x + b_{V_0}) + b_{V_1}) \\
    &= \frac {e^{V_1_i (V_0_i x + b_{V_0}_i) + b_{V_1}_i}} {\sum_j e^{V_1_j (V_0_j x + b_{V_0}_j) + b_{V_1}_j}}            
  \end{split}
  \label{eq:lr}
\end{eqnarray}

\begin{eqnarray}
  \begin{split}
    {\mathcal{L_{LR}}}(x, y) = \sum_{i=0}^{N_{X_{task}}} \log(P(Y=y^{(i)}|x^{(i)}; V_1 (V_0 x + b_{V_0}) + b_{V_1})
\end{split}
\label{eq:lr_loss}
\end{eqnarray}

\begin{wrapfigure}{r}{0.40\textwidth}
  \begin{center}
    \includegraphics[width=0.40\textwidth]{figures/figure3.png}
  \end{center}
  \caption {\textbf{Equilibrium between autoencoder and low-rank logistic regression}
  One learned decomposition component (out of 20) between the only-autoencoder
  (\textit{uppermost panel}) and only-logistic-regression
  (\textit{lowermost panel}) scenarios.
  The congruent structure (\textit{red, middle column})
  in the
  posterior cingulate cortex, posterior mid-cingulate cortex, and medial
  prefrontal cortex that is hence important in decompositions of the rest and
  task neuroimaging data. With increasing weight on the supervised learning,
  non-congruent structure emerges in the early (\textit{blue}) and
  lateral (\textit{red}) visual cortex
  (\textit{right column}). Matrix weights were z-scored.
  }
\end{wrapfigure}

\paragraph{Layer combination.}
Importantly, the optimization problem of the linear autoencoder
and the low-rank logistic regression
are linked one two levels. First, their transformation matrices mapping from
input to the latent space are identical.
    $V_0 = W_0$
We thus search for a compression of the 79,941 voxel values into $n$ latent
components that represent an optimal encoding for both
rest and task activity data.
Second, the objectives of the autoencoder and the low-rank
logistic regression are combined into a same loss function.

\begin{eqnarray}
  \begin{split}
{arg\,min}_\theta \hspace{0.2cm} {\mathcal{L}}(\theta, \lambda) = \lambda\frac{1}{N_{X_{task}}} {\mathcal{L_{LR}}}
+ (1-\lambda)\frac{1}{N_{X_{rest}}} {\mathcal{L_{AE}}} + \ell_1(\theta) + \ell_2(\theta)
  \label{eq:loss_equ}
\end{split}
\end{eqnarray}

In so doing, we search for the combined set of model parameters
$\theta=\{V_0,V_1,b_{V_0}, b_{V_1}, b_{V_0}, b_{V_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification. For parameter shrinkage, we
impose a combination of \ell_1 and \ell_1 penalties (i.e., elasticnet)
on the parameters \theta.

\paragraph{Optimization.}
The statistical learning problem was approximated in the neuroimaging data
by computing an updating the parameter derivates of the 
semi-supervised low-rank logistic regression.
The required gradients are easily obtained by using the chain rule to
backpropagate error derivatives first through the decoder network
and then through the encoder network 
As solver, we chose \textit{rmsprop} \cite{rmsprop},
a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by RMSprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to be zero.
With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by

@@

\paragraph{Implementation.}
The analyses were performed in Python.
We used \textit{nilearn} to handle
the high-dimensional neuroimaging data 
\cite{abrah14}
and
\textit{theano} for automatic
differentiation of symbolic computation graphs
\cite{bastien2012theano, bergstra2010theano}.
All Python scripts that generated the results are
accessible online for reproducibility and reuse
\url{http://github.com/banilo/nips2015}.

@@

This was evaluated by the prediction
accuracy on a validation set (20\% of the training data) at each iteration.

@@

class weights:
the right model should have high probability in only some places
